Density-dependent selection and the limits of relative fitness

### 2018-03-14 manuscript.tex
l

[Needs revising]
Selection is commonly described in terms of constant relative fitnesses. Yet when selection is strong, the ecological view of selection in density-regulated populations seems to be incompatible with constant-density relative fitnesses. Here we analyze the population ecological limits of relative fitness using a novel of density-dependent selection which contains a ``reproductive excess’’. Our model clearly distinguishes between density-dependent selection and changes in density driven by selection. These two effects are confounded in standard models of density-regulated population growth, but both are necessary, in combination with strong selection, for relative fitness to break down in populations close to demographic equilibrium. Remarkably, both effects are not sufficient: we give an example of strong selection on a density-regulating trait subject to density-dependent selection that conforms to the density-independent relative fitness description almost exactly. We reiterate the importance of reproductive excesses in many species, which allows even strong selection to have no effect on density. Our model also offers a possible alternative to relative fitness when the latter is untenable, as is likely the case far from demographic equilibrium.

@@

There are a variety of different measures of fitness. Some widely used examples are expected lifetime reproductive ratio $R_0$, intrinsic population growth rate $r$, equilibrium population density/carrying capacity (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}. In addition, ``relative fitness'' is the standard in much of evolutionary biology, particularly evolutionary genetics, where the focus is relative genotypic proportions \cite[pp. 468]{barton_2007}. The variety of fitness measures is not problematic in itself, because different measures have different uses. But it should be clear how these measures are connected to the processes of birth and death which ultimately drive selection \citep{metcalf_2007,doebeli_2017}. While such a connection is fairly clear for absolute fitness measures like $r$, relative fitness is largely divorced from population ecology. It has even been proposed that relative fitness be justified from measure theory, abandoning population biology altogether \citep{wagner_2010}.

Constant relative fitness values are typically justified as an approximation that holds when selection is sufficiently weak and stable over time (\citealt{kimura1969natural}; \citealt[pp. 277]{ewens_2004}; \citealt[Chap. 4]{charlesworth_1994}) and population density is either changing either much faster or much slower than type frequencies \citep[Fig. 9.5]{otto_2011}. Formally, these constant values are the leading terms in an expansion of a potentially complicated relative fitness function which depends on density, frequency, age structure, and so on. However, knowing that strong selection can break the relative fitness approximation in toy models like the logistic \citep{kimura1969natural}, or in highly complex age-structured models \citep[Chap. 4]{charlesworth_1994} leaves important issues unresolved. For instance, must we simply abandon relative fitness in important model systems like wild \textit{Drosophila}, where strong seasonally-alternating selection occurs in unison with ``boom-bust'' cycles in population density \citep{messer_2016,bergland_14}? And if we do abandon relative fitness, are we compelled to develop customized absolute fitness models that capture the specific ecology of the system in question? Being able to assign relative fitness values to genotypes is a powerful idealization. It is therefore important that we understand the population ecological limits of this idealization when selection is not weak and demography is not stable, both to clarify its domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.

In the absence of crowding, relative fitness simply represents differences in intrinsic population growth rate. In discrete time, the change in frequency of type $i$ is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $W_i$ is the intrinsic absolute growth factor of type $i$, and $\overline{W}=\sum_i W_i p_i$ is the population mean $W$. The change in frequency depends only the relative ratio $\frac{W_i}{\overline{W}}$, not the absolute values of the growth factors $W_i$. Thus, we can replace the $W_i$ with any set of values $w_i$ that are proportional to the $W_i$; these values are called ``relative fitnesses''. The analogous formula in continuous time is $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$, where $W_i$ is replaced by the intrinsic exponential growth rate $r_i$ \citep[pp. 26]{crow_1970}, and the ratio $\frac{W_i}{\overline{W}}$ is replaced by the difference $r_i-\overline{r}$. In the particular case that there are two types present, a wildtype $i$ and a mutant $j$ for instance, then the continuous time equation takes the familiar form
\begin{equation}
\frac{d p_i}{dt}=s p_i(1-p_i), \label{eq:canonical}
\end{equation}
where $s=r_i-r_j$ is the selection coefficient.

This simple interpretation of relative fitness breaks down when we allow for population crowding. Since crowded and uncrowded conditions are so different, $s$ will often depend on density \citep{travis_2013}. Eq.~\eqref{eq:canonical} is then no longer a complete description of selection --- we would also need to specify a model for how density is changing. Note that frequency-dependent selection does not raise similar problems; Eq.~\eqref{eq:canonical} is still a complete description of selection even if its behavior is more complicated due to $s$ depending on frequency. This issue does not arise in many population genetics models because total population density $N$ is assumed to have reached some fixed equilibrium value. The selection coefficient $s$ then abstractly parameterizes the rate at which selection changes relative frequencies, and no longer corresponds to differences in intrinsic growth rates $r$.

Yet MacArthur famously argued that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}. The classic example is the logistic model, where the type with the greatest equilibrium population density ``$K$'' excludes the others (Fig.~\ref{fig:Ksel}a). Similarly, the ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}. Differences in $R^*$ will often entail differences in equilibrium density. The Lotka-Volterra competition model also couples selection in crowded populations to density except in special cases \citep{smouse_1976,mallet_2012}. In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps. It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes driving selection (Fig.~\ref{fig:Ksel}b).

@@

In light of this, the relative fitness description has been justified in broadly two different ways for crowded populations in the strong-selection, non-equilibrium demography regime. The first is to simply assume that selection is density-independent but relax the assumption of constant $N$ by allowing density to change as a result of selective sweeps \citep[pp. 468]{barton_2007} \citep{prout_1980}. Obviously this does not address the problem that $s$ can, in reality, depend on density. Type-specific responses to density are at the center of MacArthur's argument and the density-dependent selection literature that grew out of it (e.g. \citep{roughgarden_1979}). The second justification, which primarily grew out of a controversy over Haldane's ``cost of selection'', is to appeal to the existence of a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population,kimura1969natural,nei1971fertility}. Selection can then be concentrated at the juvenile phase, uncoupling selection from population density at the adult phase unless it is so strong that the reproductive excess is depleted. Selective sweeps then need not affect density, and so the density-dependence of selection does not matter. Unfortunately this reproductive excess literature is also poorly integrated with population ecology. For instance, \cite{kimura1969natural} simply took constant $N$ as a requirement and then derived some variants of the logistic model that satisfy this requirement. \cite{nei1971fertility} proposed a model with an explicit representation of reproductive excess, but used an unusual model of competition based on pair-wise interactions which was only defined for at most two different types. As a result, the role of reproductive excesses in justifying the use of relative fitness is still largely verbal.

Here we study the population ecology of relative fitness using a novel model of density-dependent population growth based on territorial contests. We restrict our attention to asexual haploids with little or no age structure, since it is then easier to evaluate how the success or failure of the relative fitness description is tied to the underlying population ecological assumptions. Our starting point is the classic lottery model which was developed by ecologists to study competition driven by territorial contests in reef fishes and plants \citep{sale_77,chesson_1981}. The classic lottery incorporates a reproductive excess, and fitness involves a product of fertility and juvenile viability akin to the population simplest population genetic models of selection (e.g. \citealt[pp. 185]{crow_1970}). The classic lottery also assumes constant $N$, and generations can overlap. Our first task is to generalize the lottery model to create a variable-density lottery (sections ``Model'' and ``Analytical approximation of the variable-density lottery'').

Equipped with this model, we turn to the evaluation of relative fitness. [Need to write outline, will do when structure is finalized]

@@

\subsection*{Density-regulating traits at equilibrium}

>>>>

[I'm going to revise this after your next round of comments]
Selection is commonly described by assigning relative fitness values to genotypes. Yet when selection is strong, the ecological view of selection in density-regulated populations seems to be incompatible with constant-density relative fitnesses. Here we analyze the population ecological limits of relative fitness using a novel of density-dependent selection which contains a ``reproductive excess’’. Our model clearly distinguishes between density-dependent selection and changes in density driven by selection. These two effects are confounded in standard models of density-regulated population growth, but both are necessary, in combination with strong selection, for relative fitness to break down in populations close to demographic equilibrium. Remarkably, both effects are not sufficient: we give an example of strong selection on a density-regulating trait subject to density-dependent selection that conforms to the density-independent relative fitness description almost exactly. We reiterate the importance of reproductive excesses in many species, which allows even strong selection to have no effect on density. Our model also offers a possible alternative to relative fitness when the latter is untenable, as is likely the case far from demographic equilibrium.

@@

There are a variety of different measures of fitness. Some widely used examples are expected lifetime reproductive ratio $R_0$, intrinsic population growth rate $r$, equilibrium population density/carrying capacity (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}. In addition, ``relative fitness'' is the standard in much of evolutionary biology, particularly evolutionary genetics, where the focus is on relative genotypic proportions \cite[pp. 468]{barton_2007}. The variety of fitness measures is not problematic in itself, because different measures have different uses. But it should be clear how these measures are connected to the processes of birth and death which ultimately drive selection \citep{metcalf_2007,doebeli_2017}. While such a connection is fairly clear for absolute fitness measures like $r$ or $R_0$, relative fitness has only weak justification from population ecology. It has even been proposed that relative fitness be justified from measure theory, abandoning population biology altogether \citep{wagner_2010}. Given the ubiquitous use of relative fitness, it is important that we understand its population ecological basis, both to clarify its domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.

For haploids growing in discrete time steps, the change in the abundance $n_i$ of type $i$ over a time step can be expressed as $\Delta n_i=(W_i - 1)n_i$ where $W_i$ is absolute fitness. The corresponding change in frequency is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $\overline{W}=\sum_i W_i p_i$. In continuous time, the Malthusian parameter $r_i$ replaces $W_i$ and we have $\frac{d n_i}{dt}=r_in_i$ and $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$ \citep[pp. 26]{crow_1970}. In the particular case that two types are present, a mutant labeled $1$ and a wildtype labeled $2$ for instance, then the continuous time equation for the mutant takes the familiar form
\begin{equation}
\frac{d p_1}{dt}=s p_1(1-p_1), \label{eq:canonical}
\end{equation}
where $s=r_1-r_2$ is the selection coefficient. Note that in the discrete case we can replace the $W_i$ with any set of values proportional to the $W_i$ and the changes in frequency $\Delta p_i$ will be unaffected. These values, which give no information about the dynamics of total population density $N=\sum_i n_i$, are called ``relative fitnesses'' \citep[pp. 468]{barton_2007}. Similarly, the selection coefficient and frequency dynamics in Eq.~\eqref{eq:canonical} would be unaffected if we were to add an arbitrary constant to both $r_1$ and $r_2$ (these would be relative log fitnesses).

In the absence of crowding, each type grows exponentially at a constant ``instrinsic'' rate $r_i$ (for concreteness, we will primarily use Eq.~\eqref{eq:canonical} as our point of comparison but similar logic applies in the discrete time case). The selection coefficient $s$ is then a constant equal to the difference in intrinsic population growth rates. The situation is more complicated when we allow for population crowding. Since crowded and uncrowded conditions can favor quite different traits \citep{travis_2013}, the Malthusian parameters $r_i$ may respond differently to changes in population density $N$, in which case the selection coefficient will be density-dependent. Eq.~\eqref{eq:canonical} would then need to supplemented with a model for the dynamics of $N$ to describe how the allele frequencies will change over time. The density and type frequency dynamics will in general be coupled \citep{travis_2013}. Thus, in a general crowded population the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates --- a substantially more complicated problem. Note that frequency-dependent selection does not raise such a severe problem, because we still only need to model the type frequencies, and not also the ecological variable $N$.

In practice, many population genetics models assume that $N$ has reached some fixed equilibrium value, and each type is assigned a constant relative fitness. Selection is typically assumed to operate through viability (e.g. \citealt[pp. 61]{gillespie_2004}), but other important population ecological assumptions --- such as how density is regulated --- are left unspecified. Beyond being conceptually unsatisfactory \citep{doebeli_2017}, we might also doubt whether this density-independent-$s$, constant-$N$ setup is consistent with the ecology of many populations. Much of the seminal literature on density-dependent selection revolves around the connection between selection in crowded populations and the ability to keep growing at higher densities \citep{macarthur_1967,roughgarden_1979}. The classic example is the logistic model, where the type with the greatest equilibrium population density ``$K$'' excludes the others (Fig.~\ref{fig:Ksel}a). Similarly, the ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ (which will often entail a different equilibrium population density) excludes the others \citep{grover_1997}. In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps. The Lotka-Volterra competition model also couples selection in crowded populations to density except in special cases \citep{smouse_1976,mallet_2012}. It would therefore seem that the ubiquitous density-independent-$s$, constant-$N$ description of selection is incompatible with a huge class of population ecological processes driving selection (Fig.~\ref{fig:Ksel}b).

@@

In light of this, and the related difficulties posed by frequency-dependent selection, relative fitness values that are independent of type frequencies and population density are typically justified as an approximation that holds when selection is weak (\citealt{kimura1969natural}; \citealt[pp. 277]{ewens_2004}; \citealt[Chap. 4]{charlesworth_1994}) and density is changing much slower than type frequencies \citep[Fig. 9.5]{otto_2011}. Under these conditions, the selection coefficient may depend on density and type frequencies, but since these change slowly the selection coefficient may remain approximately constant over several generations. In the case of density-dependence, this time interval of constancy could include entire selective sweeps. On the other hand, it is known in both simple models like the logistic \citep{kimura1969natural}, and in complex age-structured models \citep[Chap. 4]{charlesworth_1994}, that frequency trajectories under strong density- or frequency-dependent selection can differ substantially from the constant-$s$ approximation.

Yet knowing that the constant-$s$ approximation can fail \textit{in general} when selection is strong gives only a rough picture of the domain and manner in which the constant-$s$ approximation breaks down. For instance, wild \textit{Drosophila} is an important model system in genetics, and exhibits strong seasonally-alternating selection in unison with large ``boom-bust'' density cycles \citep{messer_2016,bergland_14}. Are we compelled to switch to a more ecologically-detailed evolutionary model based on Malthusian parameters or birth/death rates in this system? And if we make this switch, how much ecological detail do we need? Perhaps the strongest justification of constant relative fitness models is that they are a powerful idealization. Thus, before abandoning them it is important to have a better understanding when and how we should do so when selection is not weak, and demography is not stable.

The rough picture given by the weak-selection approximation is missing at least one key element: virtually all models of density-dependent selection do not have a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population,kimura1969natural,nei1971fertility}. Reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles --- rather different from the classic $K$-selection scenario. Although similar ideas featured prominently in   early debates on the regulation of population density (coincidentally grounded in experiments on flies; \citealt{nicholson_1954}), much of the reproductive excess literature revolves around Haldane's ``cost of selection'' \citep{haldane_1957}, and is also poorly integrated with population ecology. For instance, \cite{kimura1969natural} simply took constant $N$ as a requirement and then derived variants of the logistic model consistent with this. \cite{nei1971fertility} proposed a model with an explicit representation of reproductive excess, but used an unusual model of competition based on pair-wise interactions which was only defined for at most two different types. As a result, the role of reproductive excesses in justifying the use of constant relative finesses is largely verbal.

Here we study the population ecology of relative fitness using a novel model of density-dependent population growth based on territorial contests. We restrict our attention to asexual haploids with little or no age structure, since it is then easier to evaluate how the success or failure of the constant-$s$ description is tied to the underlying population ecological assumptions. Our starting point is the classic lottery model which was developed by ecologists to study competition driven by territorial contests in reef fishes and plants \citep{sale_77,chesson_1981}. The classic lottery incorporates a reproductive excess, and fitness involves a product of fertility and juvenile viability akin to standard population genetic models of selection (e.g. \citealt[pp. 185]{crow_1970}). The classic lottery also assumes constant $N$, and generations can overlap. Our first task is to generalize the lottery model to create a variable-density lottery (sections ``Model'' and ``Analytical approximation of the variable-density lottery'').

Equipped with our new variable-density lottery model, we outline its selection properties (section ``Density-dependent selection in the variable-density lottery''). Next, we discuss
selection on the ability to contest territories, which behaves like a pure constant-$N$, relative fitness trait, and discuss how this fits with the classical notion of $K$-selection  (section ''The response of density to selection; $c$-selection versus $K$-selection''). We then consider selection on density-regulating traits (section ``Density-regulating traits and the threat of strong selection''), and conclude by contrasting the classical density-dependent selection literature with our results (``Discussion'').

@@

\subsection*{Density-regulating traits and the threat of strong selection}
